---
title: "Mini Data-Analysis Deliverable 1"
output: github_document
---
# Introduction
This is my very first data analysis project, and it's also a deliverable for Mini-Data Analysis Milestone 1 in STAT 545A. 

In this deliverable, I selected a dataset after evaluating several options, created a subdataset, made modifications for further research, and examined the relationships between certain variables. Based on this analysis, I have formulated four questions that I would like to explore further in Milestone 2.

For the sake of convenient grading, I have kept the relevant guidance and questions.

# Task0: Load needed packages 
1.  Install the [`datateachr`](https://github.com/UBC-MDS/datateachr) package by typing the following into **R terminal**:

<!-- -->

    install.packages("devtools")
    devtools::install_github("UBC-MDS/datateachr")

2.  Load the packages below.

```{r}
library(datateachr)
library(tidyverse)
```

# Task 1: Choose a dataset 

The `datateachr` package by Hayley Boyce and Jordan Bourak currently composed of 7 semi-tidy datasets for educational purposes. Here is a brief description of each dataset:

-   *apt_buildings*: Acquired courtesy of The City of Toronto's Open Data Portal. It currently has 3455 rows and 37 columns.

-   *building_permits*: Acquired courtesy of The City of Vancouver's Open Data Portal. It currently has 20680 rows and 14 columns.

-   *cancer_sample*: Acquired courtesy of UCI Machine Learning Repository. It currently has 569 rows and 32 columns.

-   *flow_sample*: Acquired courtesy of The Government of Canada's Historical Hydrometric Database. It currently has 218 rows and 7 columns.

-   *parking_meters*: Acquired courtesy of The City of Vancouver's Open Data Portal. It currently has 10032 rows and 22 columns.

-   *steam_games*: Acquired courtesy of Kaggle. It currently has 40833 rows and 21 columns.

-   *vancouver_trees*: Acquired courtesy of The City of Vancouver's Open Data Portal. It currently has 146611 rows and 20 columns.

---

1.1 **(1 point)** Out of the 7 datasets available in the `datateachr` package, choose **4** that appeal to you based on their description. Write your choices below:

<!-------------------------- Start your work below ---------------------------->

1: apt_buildings
2: cancer_sample
3: flow_sample
4: vancouver_trees

<!----------------------------------------------------------------------------->

1.2 **(6 points)** One way to narrowing down your selection is to *explore* the datasets. Use your knowledge of dplyr to find out at least *3* attributes about each of these datasets (an attribute is something such as number of rows, variables, class type...). The goal here is to have an idea of *what the data looks like*.

<!-------------------------- Start your work below ---------------------------->
By running the following code, it can be observed that the *apt_buildings* dataset is of type *data_frame*, which can be coerced into a *tibble*. The data set consists of **37** variables(Columns) and **3455** records(Rows). And most variables are of type *character*.
```{r}
class(apt_buildings)
glimpse(apt_buildings)
```
By running the following code, it can be observed that the *cancer_sample* data set is of type *data_frame*, which can be coerced into a *tibble*. The data set consists of **32** variables(Columns) and **569** records(Rows). And most variables are of type *double*.
```{r}
class(cancer_sample)
glimpse(cancer_sample)
```
By running the following code, it can be observed that the *flow_sample* data set is of type *data_frame*, which can be coerced into a *tibbles*. The data set consists of **7** variables(Columns) and **218** records(Rows).
```{r}
class(flow_sample)
glimpse(flow_sample)
```
By running the following code, it can be observed that the *vancouver_trees* data set is of type *data_frame*, which can be coerced into a *tibble*. The data set consists of **20** variables(Columns) and **146,611** records(Rows).
```{r}
class(vancouver_trees)
glimpse(vancouver_trees)
```
<!----------------------------------------------------------------------------->

1.3 **(1 point)** Now that you've explored the 4 datasets that you were initially most interested in, let's narrow it down to 1. What lead you to choose this one? Briefly explain your choice below.

<!-------------------------- Start your work below ---------------------------->
I choose **flow_sample** dataset. 

In general, it has the fewest variables, only 7, which is more suitable for beginners to do data analysis. Additionally, each variable has a clear and straightforward meaning, and is primarily in numeric format, making it convenient for statistical analysis.
<!----------------------------------------------------------------------------->

1.4 **(2 points)** Time for a  final decision! Going back to the beginning, it's important to have an *end goal* in mind. 

<!-------------------------- Start your work below ---------------------------->
My *end goal* is to figure out **temporal variation in flows**, exploring the relationship between **flow** and other temporal factors such as year, season and date.
<!----------------------------------------------------------------------------->

# Task 2: Exploring the dataset 

2.1 **(12 points)** Complete *4 out of the following 8 exercises* to dive deeper into your data. 

1.  Plot the distribution of a numeric variable.
2.  Create a new variable based on other variables in your data (only if it makes sense)
3.  Investigate how many missing values there are per variable. Can you find a way to plot this?
4.  Explore the relationship between 2 variables in a plot.
5.  Filter observations in your data according to your own criteria. Think of what you'd like to explore - again, if this was the `titanic` dataset, I may want to narrow my search down to passengers born in a particular year...
6.  Use a boxplot to look at the frequency of different observations within a single variable. You can do this for more than one variable if you wish!
7.  Make a new tibble with a subset of your data, with variables and observations that you are interested in exploring.
8.  Use a density plot to explore any of your variables (that are suitable for this type of plot).

2.2 **(4 points)** For each of the 4 exercises that you complete, provide a *brief explanation* of why you chose that exercise in relation to your data (in other words, why does it make sense to do that?), and sufficient comments for a reader to understand your reasoning and code.

<!-------------------------- Start your work below ---------------------------->
1.  Create a new variable based on other variables in your data (only if it makes sense)

I created a new variable *season* based on *month* to research on seasonal changes in flows later.
```{r}
# create a new data object named season_flow_sample and make modifications to it
season_flow_sample <- flow_sample %>% mutate(season = cut(month, breaks = c(0, 2, 5, 8, 11, 12), 
                      labels = c("Winter", "Spring", "Summer", "Fall", "Winter")))
head(season_flow_sample)
```
2.  Make a new tibble with a subset of your data, with variables and observations that you are interested in exploring.

By using *unique()* function, we can find that all the data is derived from one station, so we don't need to pay attention to *station* variable. 

Furthermore, I found through online research that the *sym* attribute is related to the condition of the river during sampling, which has nothing to do with temporal factors. Specifically, each element in the sym column represents one of the following: A: partial days, B: frozen conditions, E: estimated value, S: sample(s) collected on that day, NA: no other information. Therefore, I should remove it as well.
```{r}
# use unique() to see how many stations there are
unique(season_flow_sample$station_id)
# delete variables that we don't use
sub_flow_sample <- season_flow_sample %>% 
                   select(-c("station_id","sym"))
head(sub_flow_sample)
```

Besides, I noticed that there is a significant difference in *flow* values between the extreme states of maximum and minimum. It is not suitable to present the statistical results in a single plot, and these two aspects should be studied separately. Therefore, I have divided the dataset into two parts.
```{r}
# divide the dataset into two parts based on extreme_type
max_flow <- filter(sub_flow_sample, extreme_type =="maximum")
head(max_flow)
min_flow <- filter(sub_flow_sample, extreme_type =="minimum")
head(min_flow)
```

3.  Explore the relationship between 2 variables in a plot.

I'd like to explore the relationship between *year* and *flow* to see the interannual variation of maximum volume of the flow. Here, I choose line graph to present the data.
```{r}
year_flow_graph <- ggplot(max_flow, aes(year, flow)) +
                   geom_line() + 
                   geom_point(colour = "cornflowerblue") + 
                   ggtitle("Interannual Variation of Maximum Flow") +
                   xlab("Year") +
                   ylab("Flow") +
                   theme_bw()
# present the plot
print(year_flow_graph)
```

4.  Use a boxplot to look at the frequency of different observations within a single variable. You can do this for more than one variable if you wish!

I’d like to observe in which seasons the river flow reaches its maximum or minimum values, as well as the distribution of data when extreme values are reached. To achieve this, using *boxplot* is a good choice.

Because of the significant difference in values between the maximum and minimum, using *wrap* would result in the minimum values' plot appearing as a single line. Therefore, I have divided it into two separate boxplots.
```{r}
# Seasonal Maximum Flow Variation Boxplot
# Through multiple attempts, I found that the maximum flow values do not occur in winter and fall. 
# Therefore, I use only two colors to shade the box plot
season_flow_max_graph <- ggplot(max_flow, aes(x=season,y=flow, fill=season)) +
                         geom_boxplot(outlier.shape=NA, width=0.4) + 
                         geom_jitter(width=0.1, size=0.15, alpha=0.5)+
                         scale_fill_manual(values=c("lightgreen","aquamarine")) + 
                         ggtitle("Seasonal Maximum Flow Variation Boxplot")+
                         xlab("Season") + 
                         ylab("Flow") + 
                         theme_bw()
# present the plot
print(season_flow_max_graph)
```

I noticed that some *season* data are missing (NA) when the flow reaches its minimum value, which affects the graph's plotting. Therefore, I need to exclude them first. It's worth mentioning that I didn't use *na.omit()* to clean the entire dataset because some NA values in certain attributes don't affect the study of other attributes. For example, when studying the relationship between *year* and *flow*, NA values in the *month* column won't impact my research results. So, I kept these rows with NA values and would delete them when necessary.
```{r}
# Seasonal Minimum Flow Variation Boxplot
# Remove rows with NA values in the 'season' column
min_flow_omit <- min_flow[!is.na(min_flow$season),]

# Minimum flow values do not occur in summer. Therefore, I use three colors to shade the box plot
season_flow_min_graph <- ggplot(min_flow_omit, aes(x=season,y=flow, fill=season)) +
                         geom_boxplot(outlier.shape=NA, width=0.4) + 
                         geom_jitter(width=0.1, size=0.15, alpha=0.5)+
                         scale_fill_manual(values=c("snow","lightgreen","orange")) +
                         ggtitle("Seasonal Minimum Flow Variation Boxplot")+
                         xlab("Season") + 
                         ylab("Flow") + 
                         theme_bw()
# present the plot
print(season_flow_min_graph)
```


<!----------------------------------------------------------------------------->

# Task 3: Choose research questions 

**(4 points)** So far, you have chosen a dataset and gotten familiar with it through exploring the data. You have also brainstormed one research question that interested you (Task 1.4).  Now it's time to pick 4 research questions that you would like to explore in Milestone 2! Write the 4 questions and any additional comments below. 

<!--- *****START HERE***** --->
1. Explore the interannual variation of the river flow. Is there an upward or downward trend?
2. Does the interannual variation of the river flow exhibit cyclical changes? Do the maximum and minimum values fluctuate synchronously?
3. Study the relationship between seasons and river flow extremes. In which season does the flow generally reach its maximum or minimum? How does the distribution of extremes vary across different seasons?
4. Is the interannual variation of flow extremes related to the season in which they occur each year?



